A closer look at random and fixed effects panel regression in structural equation modeling using \texttt{lavaan}

### 2021-01-14 article_format/article_format_revision.tex
Started major changes for resubmission

\articletype{}

\title{A closer look at fixed effects regression in structural equation
modeling using \texttt{lavaan}}

@@

\begin{abstract}
This article provides an in-depth look at fixed effects regression in
the structural equation modeling (SEM) framework, specifically the
application of fixed effects in the \texttt{lavaan} package for
\texttt{R}. It is meant as a applied guide for researchers, covering the
underlying model specification, syntax, and summary output. Online
supplementary materials further discuss various common extensions to the
basic fixed-effect model, demonstrating how to relax model assumptions,
deal with measurement error in both the dependent and independent
variables, and include time-invariant predictors in a type of hybrid
fixed-/ random effects model.
\end{abstract}

\begin{keywords}
Fixed effects, structural equation modeling, lavaan, R, panel analysis
\end{keywords}

@@

Several years ago, \citet{Curran2011} reflected positively on the
growing use of panel studies in empirical social research. Some of the
strengths of panel data are well-known, e.g., the ability to establish
temporal precedence, increased statistical power and the reduction of
potential alternative models. However, perhaps the greatest strength of
panel data is that they allow for a more rigorous testing of substantive
theories. Panel data, i.e., repeated measures of the same observed units
(people, schools, firms, countries, etc.), allow researchers to
decompose the error term into a part that stays constant within units
and the part that changes over time. The part that does not change over
time can be seen as the combined effect of all time-invariant influences
(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
effects (FE) regression involves controlling for these time-invariant
influences via a number of various methods. It thus accounts for a
likely and common source of bias.

Structural equation modeling (SEM) is a popular regression framework.
One of its main strengths is its flexibility. Not only can complex
causal structures with multiple dependent variables be tested
simultaneously, but in longitudinal (and, more generally, hierarchical)
studies both time-varying and invariant predictors can be included, and
effects can easily be allowed to vary over time. Thus researchers can
allow for and study effects that increase or fade over time, or that
appear only in specific periods. Beyond that, with the use of latent
variables, SEM provides a way to deal with measurement error and get
closer to the true underlying constructs of interest.

There are a number of articles describing basic concept of panel model
regression, and FE regression in SEM
\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
intended as a \textit{practical guide} for researchers looking for
in-depth help with specifying FE models in SEM. It focuses on the
\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
software currently available (in terms of features like alignment,
latent variable interactions, for example), the \texttt{lavaan} package
has many benefits. First, like \texttt{R} it is open source and
completely free. For researchers dipping their toes into SEM, there is
no financial barrier to try, and no risk if they decide it is not for
them. Second, the implementation of \texttt{lavaan} in the larger
\texttt{R} environment is an enormous advantage. Instead of poring over
reams of plain text, copying out coefficients by hand, every part of the
\texttt{lavaan} output is available as an object. This means that all
aspects of the model, from fit indices, to coefficients and standard
errors, to the model matrices, can be accessed and easily integrated
into tables and plots. Furthermore, \texttt{R} can be used for a great
deal of applications. It can be used to manage and manipulate as well as
simulate data, perform symbolic algebra, run more traditional analyses
(e.g., multiple regression, logistic regression, principal component
analysis), etc. Once one is comfortable using \texttt{R}, there is no
longer any need to switch between different software for data
preparation and analysis.

The following article outlines the basic idea of panel regression, the
particularities of panel regression SEM, and shows its implementation in
\texttt{lavaan}. The focus on FE panel regression stems from the fact
that the assumptions needed to justify other common models, e.g., random
effects, are often implausible. As we will see, the practical
differences between the two are, in any event, very small. Using
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
data}, it demonstrates and annotates the code for the most basic FE
model and provides an overview of the summary output. One of the main
strengths of panel SEM compared to the more traditional methods of panel
analysis is its flexibility. Therefore, a number of potential extensions
to the basic model, including relaxing various assumptions, dealing with
measurement error in both the independent and dependent variables, as
well as the inclusion of time-invariant predictors in the form of a
hybrid fixed-/ random effects model, are shown in detail in the form of
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials}.

\hypertarget{panel}{%
\section{Panel models}\label{panel}}

To begin, let us start by reviewing a general panel model
\citep{Bollen2010}, also referred to as the `unobserved effects model'
\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials} when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align} where \(y_{it}\) is the dependent variable for unit
\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
is a \(1 \times K\) vector of time-varying covariates (which could
include a constant) linked to the dependent variable by the
\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
a \(1 \times M\) vector of time-invariant covariates linked to the
dependent variable by the \(M \times 1\) vector of coefficients in
\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
unobserved time-constant variables affecting the dependent variable and
\(\varepsilon_{it}\) is the idiosyncratic error.

@@

Somewhat unintuitively, the modern approach to FE regression treats the
unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
than a fixed parameter to be estimated for each of the cross sectional
units \citep{Wooldridge2002}. And in fact, this view carries over to the
way in which FE regression is applied in SEM. So, in order to help
facilitate the discussion on the application of FE-SEM later on, it
makes sense to spend some time discussing the basic idea of random
effects regression (again, as we will see, the practical difference
between the two models is very small).

@@

Estimation of the RE model can be done using feasible generalized least
squares (GLS) in which the two unknowns in \(\bm{\Omega}\),
\(\sigma_{\alpha}^{2}\) and \(\sigma_{\nu}^{2}\), are first estimated
using pooled ordinary least squares (pooled OLS or POLS),
where\footnote{Normally a degrees-of-freedom correction is applied by
  subtracting off the number of independent variables that is negligible
  in large N samples. It is ignored here for the sake of simplicity.}
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align} and
\(\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}\),
\(\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}\) and
\(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
Then, the coefficients are estimated using those estimates in the
variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align} where
\(\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}\) and \(\bm{X}\)
is \(NT \times K\) and \(\bm{Z}\) is \(NT \times M\) and \(\bm{y}\) is
\(NT \times 1\).

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{a =}\ErrorTok{\textasciitilde{}}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
\end{Highlighting}
\end{Shaded}

@@

Second, we regress the dependent variable on the independent variable
using the \texttt{\textasciitilde{}} regression operator. With stacked,
long-format data, only one regression coefficient is estimated over all
observed timepoints. To have our FE-SEM model mimic this behaviour, we
need to constrain the the estimated coefficient to equal over time. We
do so by adding the same label to the regression coefficient at every
time point. We will use the label \texttt{b} (this label was chosen
arbitrarily, we could have used any letter or string of characters) and
have it act as an equality constraint for the regression coefficient of
interest \(\beta\):

@@

Now, we run the FE-SEM in \texttt{lavaan}.

>>>>

\articletype{TEACHER'S CORNER}

\title{A closer look at random and fixed effects panel regression in
structural equation modeling using \texttt{lavaan}}

@@

\begin{abstract}
This article provides an in-depth look at random and fixed effects panel
regression in the structural equation modeling (SEM) framework,
specifically the application of fixed effects in the \texttt{lavaan}
package for \texttt{R}. It is meant as a applied guide for researchers,
covering the underlying model specification, syntax, and summary output.
\end{abstract}

\begin{keywords}
Fixed effects, structural equation modeling, lavaan, R, panel
regression, longitudinal data
\end{keywords}

@@

Several years ago, \citet{Curran2011} reflected positively on the
growing use of panel studies in empirical social research. Some of the
strengths of panel data are well-known, e.g., the ability to establish
temporal precedence, increased statistical power and the reduction of
potential alternative models. However, perhaps the greatest strength of
panel data is that they allow for a more rigorous testing of substantive
theories. Panel data, i.e., repeated measures of the same observed units
(people, schools, firms, countries, etc.), allow researchers to
decompose the error term into a part that stays constant within units
and the part that changes over time. The part that does not change over
time can be seen as the combined effect of all time-invariant influences
(e.g., sex, date of birth, nationality) on the dependent variable.
Random effects (RE) and fixed effects (FE) regression involve accounting
for these often unobserved time-invariant influences via a number of
methods. In the case of RE models, it is assumed that the stable
characteristics are unrelated to the other model covariates. In that
case, the stable characteristics, or individual effects as they are
often referred to as, will not affect point estimates but are a source
of error variance that, if ignored, lead to larger standard errors and
increase the likelihood of type II errors (failure to reject a false
null hypothesis). FE models are useful when the individual effects are
expected to be related to one or more of the model covariates. In that
case, these unobserved stable characteristics will act as confounders
and lead to biased point estimates. In controlling for these individual
effects, FE regression thus accounts for a likely and common source of
bias.

Structural equation modeling (SEM) is a popular regression framework.
One of its main strengths is its flexibility. Not only can complex
causal structures with multiple dependent variables be tested
simultaneously, but in longitudinal (and, more generally, hierarchical)
studies both time-varying and invariant predictors can be included, and
effects can easily be allowed to vary over time. Thus researchers can
allow for and study effects that increase or fade over time, or that
appear only in specific periods. Beyond that, with the use of latent
variables, SEM provides a way to deal with measurement error and get
closer to the theoretical constructs of interest.

There are a number of articles describing basic concept of panel model
regression, including RE and FE regression in SEM
\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
intended as a \textit{practical guide} for researchers looking for help
specifying panel regression models in SEM. It assumes some basic
knowledge of SEM (i.e., minimizing the difference between observed and
model-implied covariance matrices and mean vectors, maximum likelihood
estimation, etc.);\footnote{For an introduction to SEM,
  \citet{Bollen1989} is a classic for good reason, and
  \citet{Ferron2007} lay out exactly how SEM works in one easy to follow
  article.} perhaps the reader is an experienced SEM-user looking for a
quick guide to estimating the static panel models discussed here; or,
perhaps the reader is new to SEM, potentially because an adviser or
superior has suggested it to them. For the latter case, this article
reviews the assumptions associated with the more tradititonal least
squares-based approach to RE and FE models, so that the implementation
in SEM is more easily to follow. The article focuses on the
\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
software currently available (in terms of features like alignment,
latent variable interactions, for example), the \texttt{lavaan} package
has many benefits. First, like \texttt{R} it is open source and
completely free. For researchers dipping their toes into SEM, there is
no financial barrier to try, and no risk if they decide it is not for
them. Second, the implementation of \texttt{lavaan} in the larger
\texttt{R} environment is an enormous advantage. Instead of poring over
reams of plain text, copying out coefficients by hand, every part of the
\texttt{lavaan} output is available as an object. This means that all
aspects of the model, from fit indices, to coefficients and standard
errors, to the model matrices, can be accessed and easily integrated
into tables and plots. Furthermore, \texttt{R} can be used for a great
deal of applications. It can be used to manage and manipulate as well as
simulate data, perform symbolic algebra, run more traditional analyses
(e.g., multiple regression, logistic regression, principal component
analysis), etc. Once one is comfortable using \texttt{R}, there is no
longer any need to switch between different software for data
preparation and analysis.

The following article outlines the basic idea of panel regression, the
particularities of panel regression SEM, and shows its implementation in
\texttt{lavaan}. Using
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
data}, it demonstrates and annotates the code for the most basic FE
model and provides an overview of the summary output. One of the main
strengths of panel SEM compared to the more traditional methods of panel
analysis is its flexibility. Therefore, a number of potential extensions
to the basic model, including relaxing various assumptions, dealing with
measurement error in both the independent and dependent variables, as
well as the inclusion of time-invariant predictors in the form of a
hybrid fixed-/ random effects model, are shown in detail in the form of
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials}.

\hypertarget{panel}{%
\section{Random and Fixed Effects Panel models}\label{panel}}

It is typically the case that the values for a given unit on a variable
at one point in time will tend to tell us something about that unit's
values at another point in time. There are two main explanations for
this `empirical regularity'
\citetext{\citealp{Heckman1981}; \citealp[p.~261]{Hsiao2014}; \citealp{Bruederl2015}; \citealp{Bianconcini2018}}.
First, it could be that an experience at one point in time has a
causal\footnote{Here the word causal is used to mean `non-spurious'.}
effect on future experiences. In other words, experiencing an event
could change the probability of the same or a similar event taking place
in the future. For example, if employment increases wages, then the
incentive to continue working should increase over time, thereby making
it more likely that someone who was employed at one point in time will
continue to be employed in the future \citep{Heckman1981}. When past
experiences causally impact future events, the empirical regularity is
referred to as `state dependence'. So-called `dynamic' panel models that
include the lagged dependent variable in the equation for the current
dependent variable, like autoregressive and autoregressive cross-lagged
models, are examples of panel models that account for state dependence.
The second explanation is that correlations over time are due to stable
unobservables, like sex, place of birth, motivation, ability or
personality characteristics.\footnote{From here on out the units of
  interest are assumed to be individuals, but other units of observation
  like schools, companies, countries, etc., are possible.} In other
words, stable unit-specific characteristics might predispose individuals
to experiencing events with a certain likelihood over time. For example,
stable characteristics like an individual's sex and motivation could be
part of the reason why some individuals tend to be continuously
employed, while others experience spells of unemployment, or are
habitually unemployed. This second source of empirical regularity is
referred to as individual heterogeneity or, more generally, unobserved
heterogeneity \citep{Wooldridge2012}. So-called `static' panel models
(that do not include the lagged dependent variable in the equation for
the current one) focus on this second source of empirical regularity.

The random and fixed effects models that will be discussed here are
examples of models that attempt to account for unobserved heterogeneity
as a source of empirical regularity. In the case of both the random and
fixed effects models, it is assumed that there exist stable individual
characteristics that affect the dependent variable of interest over
time. Failure to account for this source of stable variance will reduce
model fit. If any of the stable characteristics are related to the model
covariates (a plausible scenario in many cases where the independent
variable(s) of interest are not fixed), then they must also be
considered confounders, and failure to account for them will further
lead to biased point estimates.\footnote{Whether or not the stable
  individual characteristics are related to the model covariates is the
  modern differentiator between random effects, where the individual
  effects are assumed independent of the covariates, and fixed effects,
  where the individual effects are assumed to be related to one or more
  of them, see \citet{Wooldridge2012}, p.~251. This can be a cause for
  confusion for those who are more familiar with the mixed or multilevel
  model framework. There, `fixed' effects are those coefficients that
  apply to the entire sample, `random' effects are the variance
  components for the higher level grouping variables, see for example
  \citet{Hox2010}; \citet{Bates2015}; \citet{R-lme4}.}

In this section, the basic random and fixed effects models, as well as
their implementation in SEM, will be discussed. Later on in the article,
it will be discussed how to extend the model, by relaxing some
assumptions and introducing other time-invariant predictors in a type of
`hybrid' random and fixed effects model. The outlook will describe
further extensions, such as the inclusion of lagged dependent variable
for a dynamic panel model with individual effects.\\
Let us begin with a simple linear `unobserved effects model'
\citep{Wooldridge2012, R-plm_a} \begin{align}
y_{it} & = \beta_{0} + \beta_{1} x_{it} + \alpha_{i} + \varepsilon_{it}, \ i = 1, \ldots, N, \ t = 1, \ldots, T \label{eq:panel-model}
\end{align} where \(y_{it}\) is the dependent variable for unit \(i\) at
time \(t\) and \(x_{it}\) is a single covariate with associated scalar
coefficient \(\beta_{1}\). \(\alpha_{i}\) represents the combined effect
of all unobserved time-constant variables affecting the dependent
variable and \(\varepsilon_{it}\) is the idiosyncratic error. This model
could of course be extended by turning \(x_{it}\) into a vector of
covariates
\(\bm{x}_{it} = (x_{1_{it}}, x_{2_{it}}, \ldots, x_{k_{it}})\), or by
introducing time-invariant covariates,
\(\bm{z}_{i} = (z_{1_{i}}, z_{2_{i}}, \ldots, z_{m_{i}})\), but we want
to first focus on assumptions and mechanics before moving on to
discussing possibilities for extending the model.

Equation \eqref{eq:panel-model} is often written in matrix form for unit
\(i\) as \begin{align}
\bm{y}_{i} & = \bm{\iota}_{T}\beta_{0} + \bm{x}_{i}\beta_{1} + \bm{\iota}_{T} \alpha_{i} + \bm{\varepsilon}_{i}, \ i = 1, \ldots, N \label{eq:matrix-panel-model}
\end{align} where
\(\bm{y}_{i} = (y_{i1}, y_{i2}, \ldots, y_{iT})^{\intercal}\) and
\(\bm{x}_{i} = (x_{i1}, x_{i2}, \ldots, x_{iT})^{\intercal}\) are
\(T \times 1\) vectors of each of the \(t\) observations for unit \(i\),
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones and
\(\bm{\varepsilon}_{i} = (\varepsilon_{i1}, \varepsilon_{i2}, \ldots, \varepsilon_{iT})^{\intercal}\)
is likewise a \(T \times 1\) vector of errors. It may be helpful to keep
this Equation \eqref{eq:matrix-panel-model} in mind later when we
restate the model for estimation using SEM. There, we will take the
long-format data and transform it into wide-format.

\hypertarget{assumptions-in-least-squares-based-approaches}{%
\subsection{Assumptions in least squares-based
approaches}\label{assumptions-in-least-squares-based-approaches}}

One of the main strengths of SEM is the flexibility it affords. Nearly
every part of the model is specified `by hand', so to speak (although
modern SEM software has come a long way from, say, \texttt{EQS}
\citep{Bentler2006}, where the errors and disturbances had to be
mentioned in the syntax and now comes with a variety of default settings
for ease of use), which means model assumptions are conveyed directly,
by the user specifying paths and correlations and setting other
relations to zero. Common least squares-based statistical packages
\citep[like the \texttt{plm} package in \texttt{R},][]{R-plm_a} require
less input from the user (usually it is enough to input the model
variables and specify some top-level options). The assumptions in the
least squares-based models are widely known
\citep{Wooldridge2012, Bruederl2015, Greene2012, Angrist2009} but do not
come into play prominently while specifying the model. So here we will
discuss the `typical' assumptions associated with least squares-based
random and fixed effects models. Later, when we examine the
implementation in SEM, we will then be able to relate back to this
discussion.\footnote{What constitutes a `typical' least squares-based
  approach to these panel regression models is likely debatable. Here,
  this refers broadly to the assumptions discussed in the literature and
  default settings one encounters in statistical packages intended for
  this purpose.}

We want to relate the SEM-based approach to random and fixed effects to
the least squares-based approach which is likely more widely known and
more straightforward to implement, i.e., statistical packages for static
panel regression models exist (e.g., the \texttt{plm} package in
\texttt{R}, see \citet{R-plm_a}), and typically only require the user to
enter the model variables and specify some top-level options (random
effects vs.~fixed effects, individual vs.~individual and time effects,
etc.). The model assumptions for random and fixed effects models are
widely known
\citep{Wooldridge2012, Bruederl2015, Greene2012, Angrist2009} but are
not prominent in th

The least squares-based approach to such a panel model typically makes
the following base assumptions:\footnote{What constitutes a `typical'
  least squares-based approach to these panel regression models is
  likely debatable. Here, this refers broadly to the default assumptions
  and settings one encounters in statistical packages intended for this
  purpose, like the \texttt{plm} package in \texttt{R}, see
  \citet{R-plm_a}.} first, the mean function is linear, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}(y_{it} | x_{it}, \alpha_{i}) = \beta_{0} + \beta_{1} x_{it} + \alpha_{i}\),
and second, that the cross-sectional observations are independent, which
is assured by random sampling \citep{Wooldridge2012}. Further, and just
as with any other regression model, the idiosyncratic error is assumed
to be mean independent of the model covariates and the individual
effects, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it} | x_{it}, \alpha_{i}) = \mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it}) = 0\).\footnote{If
  there is an overall intercept, \(\beta_{0}\), then it is safe to
  assume the unconditional expectation of the idiosyncratic error is
  zero, i.e., \(\mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it}) = 0\).}
The mean independence assumption also implies the assumption of a zero
correlation between the idiosyncratic error and the model covariates,
i.e., \(\mathop{\mathrm{\mathbb{E}}}(x_{it}\varepsilon_{it}) = 0\).

The (modern) difference between random and fixed effects boils down to
the assumption of the relatedness of the model covariates and the
individual effects. The random effects model assumes the model
covariates are unrelated to the individual effects, i.e., in this case
\(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i} | x_{it}) = \mathop{\mathrm{\mathbb{E}}}(\alpha_{i})\)
and thus \(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i}x_{it}) = 0\). As
such, we could put \(\alpha_{i}\) into the error term to form the
composite error, \(\nu_{it} = \alpha_{i} + \varepsilon_{it}\), and the
point estimates of the coefficients of interest, here \(\beta_{1}\),
would be consistent (assuming the model is otherwise correctly
specified). The problem with putting the individual effects into the
error term is that it would induce dependency between the compostie
errors within the same individual, which goes against the independence
assumption \citep{Wooldridge2012, Schmidheiny2019}. We will discuss that
in more detail below. In the fixed effects model, one or more of the
model covariates are assumed to be related to the individual effects,
here
\(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i} | x_{it}) \ne \mathop{\mathrm{\mathbb{E}}}(\alpha_{i})\)
and thus \(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i}x_{it}) \ne 0\). Thus,
ignoring the individual heterogeneity and putting \(\alpha_{i}\) into
the error term would mean that point estimates of the coefficients of
interest would be inconsistent.

The traditional least squares-based approaches to random and fixed
effects often assume as well that the errors are homoskedastictic over
time, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it}^{2}) = \sigma_{\varepsilon_{t}}^{2} = \sigma_{\varepsilon}^{2}\).
The individual effects are assumed to have a constant partial effect on
the dependent variable over time \citep[p.~248]{Wooldridge2012}. Also,
typically least squares-based approaches stack each of the \(N\)
repeated measures per individual \(i\) and calculate one single
coefficient per variable for the entire sample. I.e., it is generally
not intended that the estimated coefficient should vary across \(t\) or
\(i\).

\hypertarget{random-and-fixed-effects-in-sem}{%
\section{Random and fixed effects in
SEM}\label{random-and-fixed-effects-in-sem}}

For panel regression in SEM, we typically convert any stacked,
long-format vectors of observations of length \(NT\) into \(T\) separate
vectors of length \(N\). In doing so, we get \(T\) individual equations
for each time point \begin{align}
\bm{y}_{t} & = \bm{\iota}_{N}\beta_{0_{t}} + \bm{x}_{t}\beta_{1_{t}} + \bm{\alpha} + \bm{\varepsilon}_{t}, \ t = 1, \ldots, T \\
\begin{bmatrix}
y_{1_{t}} \\
y_{2_{t}} \\
\vdots \\
y_{N_{t}}
\end{bmatrix} & = 
\begin{bmatrix}
1 \\
1 \\ 
\vdots \\
1
\end{bmatrix} \beta_{0_{t}} + 
\begin{bmatrix}
x_{1_{t}} \\
x_{2_{t}} \\
\vdots \\
x_{N_{t}}
\end{bmatrix} \beta_{1_{t}} + 
\begin{bmatrix}
\alpha_{1} \\
\alpha_{2} \\
\vdots \\
\alpha_{N}
\end{bmatrix} + 
\begin{bmatrix}
\varepsilon_{1_{t}} \\
\varepsilon_{2_{t}} \\
\vdots \\
\varepsilon_{N_{t}}
\end{bmatrix}
\end{align}

Let us discuss the random and fixed effects models in SEM by examining a
path diagram. Figure \ref{fig:panel-sem} shows a four-wave panel model
with individual effects. As usual, rectangles indicate observed
variables and circles indicate latent variables. One-sided errors
indicate effects and two-headed errors indicate correlations or
covariances.

We assume all of the observed variables are grand mean centered, so that
we can drop the overall intercept from the model.

\begin{figure}
\begin{center}
\resizebox{0.75\textwidth}{!}{%
\begin{tikzpicture}
% Node styles ---
\tikzstyle{man} = [rectangle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
\tikzstyle{lat} = [circle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
\tikzstyle{err} = [rectangle, minimum size = 0.5cm, draw = white!100, fill = white!100, font = \sffamily] 
\tikzstyle{pha} = [rectangle, minimum size = 0.001cm, draw = white, fill = white]
\tikzstyle{con} = [-latex, draw = black!100, font = \sffamily] 
\tikzstyle{seqex} = [latex-latex, draw = black!100, font = \sffamily, dashed]
\tikzstyle{cons} = [-latex, draw = black!100, font = \sffamily\small] 
\tikzstyle{cor} = [latex-latex, font = \sffamily] 
% Begin Figure ---
% Nodes 
\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
\node at (+8.0,+0.0) [man] (x4) {$x_{4}$};
\node at (+4.0,+0.0) [lat] (alp) {$\alpha$};
\node at (+0.0,+2.1) [man] (y1) {$y_{1}$};
\node at (+2.0,+2.1) [man] (y2) {$y_{2}$};
\node at (+6.0,+2.1) [man] (y3) {$y_{3}$};
\node at (+8.0,+2.1) [man] (y4) {$y_{4}$};
\node at (+0.0,+3.0) [err] (e1) {\footnotesize $\varepsilon_{1}$};
\node at (+2.0,+3.0) [err] (e2) {\footnotesize $\varepsilon_{2}$};
\node at (+6.0,+3.0) [err] (e3) {\footnotesize $\varepsilon_{3}$};
\node at (+8.0,+3.0) [err] (e4) {\footnotesize $\varepsilon_{4}$};
% Paths
\path (alp) edge [con] node [near start, fill=white] {1} (y1);
\path (alp) edge [con] node [midway, fill=white] {1} (y2);
\path (alp) edge [con] node [midway, fill=white] {1} (y3);
\path (alp) edge [con] node [near start, fill=white] {1} (y4);
\path (x1) edge [con] node [near start, fill=white] {$\beta$} (y1);
\path (x2) edge [con] node [near start, fill=white] {$\beta$} (y2);
\path (x3) edge [con] node [near start, fill=white] {$\beta$} (y3);
\path (x4) edge [con] node [near start, fill=white] {$\beta$} (y4);
\path (e1) edge [con] node {} (y1);
\path (e2) edge [con] node {} (y2);
\path (e3) edge [con] node {} (y3);
\path (e4) edge [con] node {} (y4);
% correlations
\path (x1) edge [cor, bend right] node {} (x2);
\path (x1) edge [seqex, bend right] node {} (alp);
\path (x1) edge [cor, bend right] node {} (x3);
\path (x1) edge [cor, bend right] node {} (x4);
\path (x2) edge [seqex, bend right] node {} (alp);
\path (x2) edge [cor, bend right] node {} (x3);
\path (x2) edge [cor, bend right] node {} (x4);
\path (x3) edge [seqex, bend left] node {} (alp);
\path (x4) edge [seqex, bend left] node {} (alp); 
\end{tikzpicture}
}
\caption{Four-wave static panel model\label{fig:panel-sem}}
\end{center}
\end{figure}

To begin, let us review a general panel model \citep{Bollen2010}, also
referred to as the `unobserved effects model'
\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials} when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align} where \(y_{it}\) is the dependent variable for unit
\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
is a \(1 \times K\) vector of time-varying covariates (which could
include a constant) linked to the dependent variable by the
\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
a \(1 \times M\) vector of time-invariant covariates linked to the
dependent variable by the \(M \times 1\) vector of coefficients in
\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
unobserved time-constant variables affecting the dependent variable and
\(\varepsilon_{it}\) is the idiosyncratic error.

@@

@@

Estimation of the RE model can be done using feasible generalized least
squares (GLS) in which the two unknowns in \(\bm{\Omega}\),
\(\sigma_{\alpha}^{2}\) and \(\sigma_{\nu}^{2}\), are first estimated
using pooled ordinary least squares (pooled OLS or POLS),
where\footnote{Normally a degrees-of-freedom correction is applied by
  subtracting off the number of independent variables that is negligible
  in large N samples. It is ignored here for the sake of simplicity.}
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align} and
\(\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}\),
\(\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}\) and
\(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
Here, \(\bm{\hat{\beta}}_{\tiny{POLS}}\) and
\(\hat{\bm{\gamma}}_{\tiny{POLS}}\) are the POLS estimates for the
coefficients of the time-varying and -invariant covariates,
respectively. Then, the coefficients are estimated using those estimates
in the variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align} where
\(\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}\) and \(\bm{X}\)
is \(NT \times K\) and \(\bm{Z}\) is \(NT \times M\) and \(\bm{y}\) is
\(NT \times 1\).

@@

@@

@@
